---
title:  "INAUSPICIOUS IMPACTS OF U.S STORMS ON HEALTH AND ECONOMY"
author: "Shubhendu Dash"
date:   "08/07/2020"
output: 
        pdf_document:   default
        html_document:  default
---

Github Repository for the Project: [Reproducible Research Project 2](https://github.com/ShubhenduDash/ReproducibleResearch_Project2)

## 1. Synopsis

Tempests and other extreme climate occasions have gigantic effect on general wellbeing and monetary issues for districts and their occupants. Some of serious occasions can cause wounds property harm and even lead to death. This investigation present which kinds of occasions are generally hurtful concerning populace wellbeing and which have the best financial outcomes. 

The objective of the task is to investigate the NOAA Storm Database and investigate the impacts of extreme climate occasions on both populace and economy.The database covers the timeframe among 1950 and November 2011. 

The accompanying examination researches which sorts of extreme climate occasions are generally destructive on:

1. Health (injuries and fatalities) 
2. Property and crops (economic consequences)

Information on the Data: [Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

## 2. Data Processing

### 2.1: Data Loading

Download the raw data file and extract the data into a dataframe and converting it into data.table.

```{r DataLoading, cache=TRUE}

# Reading packages

library("data.table")
library("ggplot2")

# Downloading data

USStormNOAA <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url = USStormNOAA, destfile = "USStormDataNOAA")

# Reading data

stormDF <- read.csv(bzfile("USStormDataNOAA"),sep = ",",header=TRUE)

# Converting data.frame to data.table

stormDT <- as.data.table(stormDF)

```

### 2.2: Examining Column Names

```{r ColumnNames, cache=TRUE}

colnames(stormDT)

```

### 2.3: Subsetting required data

Subset the dataset on the parameters of interest. Basically, we remove the columns we don't need for clarity. 

```{r DataSubsetting, cache=TRUE}

# Finding columns to remove

cols2Remove <- colnames(stormDT[, !c("EVTYPE"
  , "FATALITIES"
  , "INJURIES"
  , "PROPDMG"
  , "PROPDMGEXP"
  , "CROPDMG"
  , "CROPDMGEXP")])

# Removing columns

stormDT[, c(cols2Remove) := NULL]

# Only use data where fatalities or injuries occurred.  

stormDT <- stormDT[(EVTYPE != "?" & (INJURIES > 0 | FATALITIES > 0 | PROPDMG > 0 | CROPDMG > 0)), c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP") ]

```

### 2.4: Converting Exponent Columns into Actual Exponents instead of (-,+, H, K, etc)

Making the PROPDMGEXP and CROPDMGEXP columns cleaner so they can be used to calculate property and crop cost.

```{r CorrectingExponents, cache=TRUE}

# Change all damage exponents to uppercase.

cols <- c("PROPDMGEXP", "CROPDMGEXP")
stormDT[,  (cols) := c(lapply(.SD, toupper)), .SDcols = cols]

# Map property damage alphanumeric exponents to numeric values.

propDmgKey <-  c("\"\"" = 10^0,
                 "-" = 10^0, 
                 "+" = 10^0,
                 "0" = 10^0,
                 "1" = 10^1,
                 "2" = 10^2,
                 "3" = 10^3,
                 "4" = 10^4,
                 "5" = 10^5,
                 "6" = 10^6,
                 "7" = 10^7,
                 "8" = 10^8,
                 "9" = 10^9,
                 "H" = 10^2,
                 "K" = 10^3,
                 "M" = 10^6,
                 "B" = 10^9)

# Map crop damage alphanumeric exponents to numeric values

cropDmgKey <-  c("\"\"" = 10^0,
                "?" = 10^0, 
                "0" = 10^0,
                "K" = 10^3,
                "M" = 10^6,
                "B" = 10^9)

stormDT[, PROPDMGEXP := propDmgKey[as.character(stormDT[,PROPDMGEXP])]]
stormDT[is.na(PROPDMGEXP), PROPDMGEXP := 10^0 ]
stormDT[, CROPDMGEXP := cropDmgKey[as.character(stormDT[,CROPDMGEXP])] ]
stormDT[is.na(CROPDMGEXP), CROPDMGEXP := 10^0 ]

```

### 2.5: Making Economic Cost Columns

```{r EconomicCostColumns, cache=TRUE}

stormDT <- stormDT[, .(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, propCost = PROPDMG * PROPDMGEXP, CROPDMG, CROPDMGEXP, cropCost = CROPDMG * CROPDMGEXP)]

```

### 2.6: Calcuating Total Property and Crop Cost

```{r TotalPropertyCropCost, cache=TRUE}

totalCostDT <- stormDT[, .(propCost = sum(propCost), cropCost = sum(cropCost), Total_Cost = sum(propCost) + sum(cropCost)), by = .(EVTYPE)]
totalCostDT <- totalCostDT[order(-Total_Cost), ]
totalCostDT <- totalCostDT[1:10, ]
head(totalCostDT, 5)

```

### 2.7: Calcuating Total Fatalities and Injuries

```{r TotalFatalitiesInjuriesCalc, cache=TRUE}

totalInjuriesDT <- stormDT[, .(FATALITIES = sum(FATALITIES), INJURIES = sum(INJURIES), totals = sum(FATALITIES) + sum(INJURIES)), by = .(EVTYPE)]
totalInjuriesDT <- totalInjuriesDT[order(-FATALITIES), ]
totalInjuriesDT <- totalInjuriesDT[1:10, ]
head(totalInjuriesDT, 5)

```

## 3. Results

### 3.1: Events that are Most Harmful to Population Health

Melting data.table so that it is easier to put in bar graph format 

```{r HealthResults, cache=TRUE}

bad_stuff <- melt(totalInjuriesDT, id.vars="EVTYPE", variable.name = "bad_thing")
head(bad_stuff, 5)

```

```{r healthChart, cache=TRUE}

# Create chart

healthChart <- ggplot(bad_stuff, aes(x=reorder(EVTYPE, -value), y=value))

# Plot data as bar chart

healthChart = healthChart + geom_bar(stat="identity", aes(fill=bad_thing), position="dodge")

# Format y-axis scale and set y-axis label

healthChart = healthChart + ylab("Frequency Count") 

# Set x-axis label
healthChart = healthChart + xlab("Event Type") 

# Rotate x-axis tick labels 

healthChart = healthChart + theme(axis.text.x = element_text(angle=45, hjust=1))

# Set chart title and center it

healthChart = healthChart + ggtitle("Top 10 Events") + theme(plot.title = element_text(hjust = 0.5))
healthChart

```

### 3.2: Events that have the Greatest Economic Consequences

Melting data.table so that it is easier to put in bar graph format 

```{r EconConsequences, cache=TRUE}

econ_consequences <- melt(totalCostDT, id.vars="EVTYPE", variable.name = "Damage_Type")
head(econ_consequences, 5)

```

```{r econChart, cache=TRUE}

# Create chart

econChart <- ggplot(econ_consequences, aes(x=reorder(EVTYPE, -value), y=value))

# Plot data as bar chart

econChart = econChart + geom_bar(stat="identity", aes(fill=Damage_Type), position="dodge")

# Format y-axis scale and set y-axis label

econChart = econChart + ylab("Cost (dollars)") 

# Set x-axis label

econChart = econChart + xlab("Event Type") 

# Rotate x-axis tick labels 

econChart = econChart + theme(axis.text.x = element_text(angle=45, hjust=1))

# Set chart title and center it

econChart = econChart + ggtitle("Top 10 US Storm Events causing Economic Consequences") + theme(plot.title = element_text(hjust = 0.5))
econChart

```

## 4. Conclusion

**As should be obvious above flood has the best financial outcomes. Twister is the most unsafe to populace wellbeing in light of the fact that caused the most losses of life and wounds.**